---
title: "Analyzing Storm Data to Determine which Weather Events Causes the Greatest Harm in Terms of Population and Economic Health"
author: "Ashley Campisano"
date: "July 14, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
###Reproducible Research - Week 4 Assignment

###Assignment Synopsis

###The goals of this assignment was to first analyze the Storm Data provided by National Weather Service in the National Oceanic and Atmospheric Administration's (NOAA) database and determine which event types are the most harmful with respect to population health. The second goal of this assignment was to use the same NOAA storm data to determine across the United States, which events have the greatest economic consequences. 

###The goals of this assignment are important as they help us to better understand the impact that severe weather events can have on both the health of the population and of the economy. As time passes, weather events are increasing in severity and frequency which heightens their potential harm to the United States and world beyond.

###Through analysis, tornadoes and heat weather events cause the most damage to population health (as measured by total fatalities and injuries). Floods and Typhoon weather events cause the most damage to economic health (as measured by total crop and property damage).

###Load packages that may be needed
```{r}
library(dplyr)
```
###Data Processing

###Load the data and briefly look into the data
```{r}
if(!file.exists("/StormData.csv.bz2")){
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",destfile="./StormData.csv.bz2")
}
stormdata <- read.csv(bzfile("StormData.csv.bz2"), sep=",", header=T)

head(stormdata)
summary(stormdata)
names(stormdata)
dim(stormdata)
```
###Create Data Subset to Analyze Event Types and Resultant Harm on Population Health
```{r}
storm_event <- stormdata[ , c(8, 23:28)]
```
###This subset provides a more manageable dataset which isolates the variables that deal with measuring weather event types and resuling fatalities, injuries, etc.

###Investigate Event Types and the Resultant Harm on Population Health
```{r}
head(storm_event)
summary(storm_event)
```     
###Investigate fatalities and inuries associated with weather events
```{r}
summary(storm_event$FATALITIES)
summary(storm_event$INJURIES)
```
###It appears that most weather events do not result in injuries/fatalities but the most harmful events resulted in a max of 1700 inuries and 583 fatalities.

###Convert Crop and Property Damage into two new variables using H=10^2, K=10^3, M =10^6, and B=10^9.
```{r}
storm_event$PROPDAMAGE = 0
storm_event[storm_event$PROPDMGEXP == "H", ]$PROPDAMAGE = storm_event[storm_event$PROPDMGEXP == "H", ]$PROPDMG * 10^2
storm_event[storm_event$PROPDMGEXP == "K", ]$PROPDAMAGE = storm_event[storm_event$PROPDMGEXP == "K", ]$PROPDMG * 10^3
storm_event[storm_event$PROPDMGEXP == "M", ]$PROPDAMAGE = storm_event[storm_event$PROPDMGEXP == "M", ]$PROPDMG * 10^6
storm_event[storm_event$PROPDMGEXP == "B", ]$PROPDAMAGE = storm_event[storm_event$PROPDMGEXP == "B", ]$PROPDMG * 10^9

storm_event$CROPDAMAGE = 0
storm_event[storm_event$CROPDMGEXP == "H", ]$CROPDAMAGE = storm_event[storm_event$CROPDMGEXP == "H", ]$CROPDMG * 10^2
storm_event[storm_event$CROPDMGEXP == "K", ]$CROPDAMAGE = storm_event[storm_event$CROPDMGEXP == "K", ]$CROPDMG * 10^3
storm_event[storm_event$CROPDMGEXP == "M", ]$CROPDAMAGE = storm_event[storm_event$CROPDMGEXP == "M", ]$CROPDMG * 10^6
storm_event[storm_event$CROPDMGEXP == "B", ]$CROPDAMAGE = storm_event[storm_event$CROPDMGEXP == "B", ]$CROPDMG * 10^9
```
###Results

###1. Across the United States, which types of events are most harmful with respect to population health?

###Aggregate Injuries Data
```{r}
injuries <- aggregate(INJURIES ~ EVTYPE, data = storm_event, FUN = sum)
injuries <- injuries[order(injuries$INJURIES, decreasing = TRUE), ]
```
###10 Most Harmful Events in Terms of Inuries
```{r}
maxinjuries <- injuries[1:10, ]
print(maxinjuries)
```
###Tornadoes are the most harmful event in terms of injuries.

###Aggregate Fatalities Data
```{r}
fatalities <- aggregate(FATALITIES ~ EVTYPE, data = storm_event, FUN = sum)
fatalities <- fatalities[order(fatalities$FATALITIES, decreasing = TRUE), ]
```
###10 Most Harmful Events in Terms of Fatalities
```{r}
maxfatalities <- fatalities[1:10, ]
print(maxfatalities)
```
###Like injuries, Tornadoes are the most harmful event in terms of fatalities.

###Plotting the Most Harmful Events in Terms of Injuries and Fatalities
```{r}
par(mfrow = c(1, 2))
barplot(maxfatalities$FATALITIES, las = 3, names.arg = maxfatalities$EVTYPE, main = "Weather Events w/Most Fatalities", ylab = "Total Fatalities", col = "lightblue")
barplot(maxinjuries$INJURIES, las = 3, names.arg = maxinjuries$EVTYPE, main = "Weather Events w/Most Injuries", ylab = "Total Injuries", col = "lightblue")
```
###Per the histograms above, tornadoes and heat resulted in the most fatalities while tornadoes and wind resulted in the most inuries.

###2. Across the United States, which types of events have the greatest economic consequences?

###Aggregate Property and Economic Damage into One Variable
```{r}
economicharm <- aggregate(PROPDAMAGE + CROPDAMAGE ~ EVTYPE, storm_event, sum)
names(economicharm) = c("EVENT_TYPE", "TOTAL_DAMAGE")
economicharm <- arrange(economicharm, desc(TOTAL_DAMAGE))
economicharm <- economicharm[1:20, ]
economicharm$TOTAL_DAMAGE <- economicharm$TOTAL_DAMAGE/10^9
economicharm$EVENT_TYPE <- factor(economicharm$EVENT_TYPE, levels = economicharm$EVENT_TYPE)

head(economicharm)
```
###Plotting the Most Harmful Events in Terms of Economic Damage
```{r}
par(mfrow = c(1, 1))
barplot(economicharm$TOTAL_DAMAGE, names.arg = economicharm$EVENT_TYPE, beside = T, cex.names = 0.8, las=2, col = "darkgreen", main = "Most Harmful Weather Events by Property and Crop Damage", ylab = "Total Damage in USD (10^9)")
```
###Based on the plot above, floods and typhoons did the greatest economic harm in terms of property and crop damage.